Gaps in stellar streams as a result of globular cluster fly-bys

### 2025-01-27 paper/main.tex
$\Lambda$~CDM to $\Lambda$CDM

\abstract
  %  menntion that only one Galactic model should be sufficient to avoid the criticism
  {Thin stellar streams, such as those resulting from the tidal disruption of globular clusters, have long been known and used as probes of the gravitational potential of our Galaxy, both its visible and dark contents. In particular, the presence of under-density regions, or gaps, along these streams is commonly interpreted as being due to the close passage of dark matter sub-halos. }
 % aims heading (mandatory)
  {In this work, we investigate the perturbations induced on streams by the passage of dense stellar systems, such as globular clusters themselves, to test the possibility that they may cause the formation of gaps as well. In particular, we focus on the study of the stream of Palomar~5, a well-known globular cluster in the Galactic halo, which has particularly long tidal tails.  }
 % methods heading (mandatory)
  {For this purpose, we used a particle-test code to simulate the dynamic evolution of Palomar~5 and its tidal tails when Palomar~5 is subject to the gravitational action of the Galaxy and its whole system of globular clusters.}
 % results heading (mandatory)
  {Our study shows that the tails of Palomar~5 are strongly perturbed by the close passage of other clusters, in particular of NGC~2808, NGC~7078, NGC~104, and that these perturbations induce the formation of gaps in the tails.}
 % conclusions heading (optional), leave it empty if necessary
  {This result suggests that baryonic dense stellar systems such as globular clusters are capable of causing gaps in streams, as already demonstrated in the case of galaxy bar-induced perturbations and giant molecular clouds, and reinforce the idea that the effect of these components should be fully addressed when interpreting the morphology and density structures of observed stellar streams.}

@@

  % Beyond the Galactic scale, dark matter is essential for explaining many cosmological phenomena. For instance, the Planck mission \citep{2020A&A...641A...1P} was designed to study the Cosmic Microwave Background (CMB), whose spectrum closely follows that of a black body with a peak temperature in the millimeter wavelength range, corresponding to approximately 2.725~K. However, there are slight temperature variations on the order of one part in ten thousand. Notably, the colder regions of the CMB temperature map correlate with whole-sky gravitational lensing maps \citep{2020A&A...641A...8P}--since the photons lost energy as they escaped the gravitational wells. The $\Lambda$~Cold Dark Matter ($\Lambda$~CDM) model accounts for these observations almost perfectly, proposing that dark matter exists and that its distribution originated from stochastic quantum fluctuations within the first second of the universe led to density variations. These variations propagated as the universe expanded, shaping the large-scale structure and distribution of galaxies we observe today.

@@

  % \citet{2019Galax...7...81Z} provide a review for the different proposed mechanisms of forming dark matter haloes theoretically and how to model them practically, showcasing how different assumptions for the fundamental particle behavior of dark matter propogates to the mass and size distribution of sub haloes on the galactic scale. The authors rightly state that until particles are discovered $\Lambda$~CDM remains a hypothesis. Since the true nature of dark matter is elusive the possibilities for sub-galactic behaviour and structure of dark matter are vast. Obtaining constraints on the distirbution of dark matter subhaloes in turn will place constraints on the particle nature of dark matter. As proposed by \citet{2012ApJ...748...20C}, a most promising tool to set limits on the number and mass density of the sub-halos, are stellar streams.

@@

  \citet{2002MNRAS.332..915I} first suggested that dark matter sub-halos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that sub-halos could create gaps in stellar streams during flyby encounters, where a sub-halo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars.  \citet{2012ApJ...760...75C, 2013ApJ...768..171C} applied this idea and looked for the presence of gaps in the known Palomar~5 and GD-1 streams, concluding that the density variations found in their streams were consistent with expectations from $\Lambda$~CDM models.  \citet{2019ApJ...880...38B} provided further observational evidence for this idea, identifying a gap and a spur in the GD-1 stream that they could not explain by known objects, such as globular clusters, and which they suggested was due to the close passage of a dark matter sub-halo. Interestingly enough, as stated in \citet{2019ApJ...880...38B}, the recovered properties of this sub-halo (mass and size) were inconsistent with the $\Lambda$~CDM mass-size relationship presented in \cite{2017MNRAS.466.4974M}. The above-mentioned works are only a few of a much more extensive literature that has explored the impact of dark matter sub-halos in simulated streams \citep{2016ApJ...828L..10H, 2021MNRAS.507.1999H, 2021JCAP...10..043B, 2024arXiv240402953H, 2024arXiv241021174N}, or searched for their traces in observed streams \citep{2016MNRAS.460.2711T, 2017MNRAS.470...60E, 2020ApJ...889...70B, 2020ApJ...892L..37B}.

@@

  As already suggested in previous works which have studied the impact of baryonic structures on Palomar~5 tails \citep{2017NatAs...1..633P, 2019MNRAS.484.2009B}, it is possible that this cluster lies in a region of the phase-space which is not favorable to distinguish gaps created by dark matter sub-halos from gaps created by baryonic structures, such as the Galactic bar and giant molecular clouds. With our study, we show that close encounters with globular clusters constitute a further element that confuses a simple interpretation of the observed Palomar~5 stream gaps (if any). Other clusters and streams in the orbital energy range of Palomar~5 may suffer of the same difficulty. Going to lower orbital energies make the situation even worse, because the effect of the Galactic bar, giant molecular clouds and interactions with globular clusters becomes even more efficient. In the innermost regions of the Galaxy, where the density of dark matter sub-halos is expected to be maximal, globular cluster streams are intrinsically more difficult to find, in addition to the fact that the dynamic times in these regions become so small that the stars lost from globular clusters do not redistribute themselves for the most part into thin structures \citep[see][]{2023A&A...673A..44F}. It is thus probably only at larger distances from the Galactic center than those spanned by Palomar~5's orbit that the impact of dark matter sub-halos may become dominant, but it should be borne in mind that at these distances the number density of sub-halos also decreases. In the future, it will be interesting to apply the type of study conducted here to the whole set of clusters for which streams have been found to quantify the regime in which they are found, whether favorable to the creation of gaps from baryonic structures or not.

>>>>

\abstract
  %  menntion that only one Galactic model should be sufficient to avoid the criticism
  {Thin stellar streams, such as those resulting from the tidal disruption of globular clusters, have long been known and used as probes of the gravitational potential of our Galaxy, both its visible and dark contents. In particular, the presence of under-density regions, or gaps, along these streams is commonly interpreted as being due to the close passage of dark matter sub-halos. }
 % aims heading (mandatory)
  {In this work, we investigate the perturbations induced on streams by the passage of dense stellar systems, such as globular clusters themselves, to test the possibility that they may cause the formation of gaps as well. In particular, we focus on the study of the stream of Palomar~5, a well-known globular cluster in the Galactic halo, which has particularly long tidal tails.  }
 % methods heading (mandatory)
  {For this purpose, we used a particle-test code to simulate Palomar~5's tidal tails when subjected to the Galaxy's gravitational field plus its whole system of globular clusters.}
 % results heading (mandatory)
  {Our study shows that the tails of Palomar~5 are strongly perturbed by the close passage of other clusters, in particular of NGC~2808, NGC~7078, NGC~104, and that these perturbations induce the formation of gaps in the tails.}
 % conclusions heading (optional), leave it empty if necessary
  {This result suggests that baryonic dense stellar systems such as globular clusters are capable of causing gaps in streams, as already demonstrated in the case of galaxy bar-induced perturbations and giant molecular clouds, and reinforce the idea that the effect of these components should be fully addressed when interpreting the morphology and density structures of observed stellar streams.}

@@

  % Beyond the Galactic scale, dark matter is essential for explaining many cosmological phenomena. For instance, the Planck mission \citep{2020A&A...641A...1P} was designed to study the Cosmic Microwave Background (CMB), whose spectrum closely follows that of a black body with a peak temperature in the millimeter wavelength range, corresponding to approximately 2.725~K. However, there are slight temperature variations on the order of one part in ten thousand. Notably, the colder regions of the CMB temperature map correlate with whole-sky gravitational lensing maps \citep{2020A&A...641A...8P}--since the photons lost energy as they escaped the gravitational wells. The $\Lambda$~Cold Dark Matter ($\Lambda$CDM) model accounts for these observations almost perfectly, proposing that dark matter exists and that its distribution originated from stochastic quantum fluctuations within the first second of the universe led to density variations. These variations propagated as the universe expanded, shaping the large-scale structure and distribution of galaxies we observe today.

@@

  % \citet{2019Galax...7...81Z} provide a review for the different proposed mechanisms of forming dark matter haloes theoretically and how to model them practically, showcasing how different assumptions for the fundamental particle behavior of dark matter propogates to the mass and size distribution of sub haloes on the galactic scale. The authors rightly state that until particles are discovered $\Lambda$CDM remains a hypothesis. Since the true nature of dark matter is elusive the possibilities for sub-galactic behaviour and structure of dark matter are vast. Obtaining constraints on the distirbution of dark matter subhaloes in turn will place constraints on the particle nature of dark matter. As proposed by \citet{2012ApJ...748...20C}, a most promising tool to set limits on the number and mass density of the sub-halos, are stellar streams.

@@

  \citet{2002MNRAS.332..915I} first suggested that dark matter sub-halos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that sub-halos could create gaps in stellar streams during flyby encounters, where a sub-halo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars.  \citet{2012ApJ...760...75C, 2013ApJ...768..171C} applied this idea and looked for the presence of gaps in the known Palomar~5 and GD-1 streams, concluding that the density variations found in their streams were consistent with expectations from $\Lambda$CDM models.  \citet{2019ApJ...880...38B} provided further observational evidence for this idea, identifying a gap and a spur in the GD-1 stream that they could not explain by known objects, such as globular clusters, and which they suggested was due to the close passage of a dark matter sub-halo. Interestingly enough, as stated in \citet{2019ApJ...880...38B}, the recovered properties of this sub-halo (mass and size) were inconsistent with the $\Lambda$CDM mass-size relationship presented in \cite{2017MNRAS.466.4974M}. The above-mentioned works are only a few of a much more extensive literature that has explored the impact of dark matter sub-halos in simulated streams \citep{2016ApJ...828L..10H, 2021MNRAS.507.1999H, 2021JCAP...10..043B, 2024arXiv240402953H, 2024arXiv241021174N}, or searched for their traces in observed streams \citep{2016MNRAS.460.2711T, 2017MNRAS.470...60E, 2020ApJ...889...70B, 2020ApJ...892L..37B}.

@@

  As already suggested in previous works which have studied the impact of baryonic structures on Palomar~5 tails \citep{2017NatAs...1..633P, 2019MNRAS.484.2009B}, it is possible that this cluster lies in a region of the phase-space which is not favorable to distinguish gaps created by dark matter sub-halos from gaps created by baryonic structures, such as the Galactic bar and giant molecular clouds. With our study, we show that close encounters with globular clusters constitute a further element that confuses a simple interpretation of the observed Palomar~5 stream gaps (if any). Other clusters and streams in the orbital energy range of Palomar~5 may suffer of the same difficulty. Going to lower orbital energies make the situation even worse, because the effect of the Galactic bar, giant molecular clouds and interactions with globular clusters becomes even more efficient. In the innermost regions of the Galaxy, where the density of dark matter sub-halos is expected to be maximal, globular cluster streams are intrinsically more difficult to find, in addition to the fact that the dynamical times in these regions become so small that the stars lost from globular clusters do not redistribute themselves for the most part into thin structures \citep[see][]{2023A&A...673A..44F}. It is thus probably only at larger distances from the Galactic center than those spanned by Palomar~5's orbit that the impact of dark matter sub-halos may become dominant, but it should be borne in mind that at these distances the number density of sub-halos also decreases. In the future, it will be interesting to apply the type of study conducted here to the whole set of clusters for which streams have been found to quantify the regime in which they are found, whether favorable to the creation of gaps from baryonic structures or not.
